Experience Report: Haskell in Computational Biology

### 2012-07-02 mrfy_experience_report.tex
good page breaks; now let's check content

\category{D.1.1}{Applicative (Functional) Programming}{}
\category{J.3}{Biology and genetics}{}
% \category{CR-number}{subcategory}{third-level}
% 
% \terms
% term1, term2
% 
\keywords
memoization, stochastic search, parallel strategies,  QuickCheck, remote homology detection

@@

To~account for beta strands, computational biologists add
a term to Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on whatever amino acid~$x'$  is in column~${\pairedwith i}$.
If~$x'$ is located in position~$i'$ in the query sequence, Viterbi's
equations must be altered so that, for example
$V_{j}^{\prime M}(i)$ depends not only on
$V_{j-1}^{\prime M}(i-1)$ but also on
$V_{\pairedwith j}^{\prime M}(i')$.\remark{I can't get an equation I
believe in. ---NR I think it's fine, unless we want to introduce
notation to distinguish paired amino acids from paired
model nodes --ND}
The distance between $j$~and~$\pairedwith j$ can be as small as a few
columns or as large as a few hundreds of columns.
Because $V_j^{\prime M}(i)$~depends not only on nearby values but also on
$V_{\pairedwith j}^{\prime M}(i')$,
dynamic programming cannot compute the maximum likelihood quickly 
\cite{Menke:2010ti,Daniels:2012}.

The new terms in Viterbi's equations are accompanied by a new model.
Within a beta strand, amino acids are not inserted or deleted, so
a bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}.

MRFy treats the beta 
strands in the model
 as ``beads'' which can slide along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the remaining parts of
the query sequence, which appear between and around beta strands,
can be computed quickly and exactly 
using Viterbi's algorithm.
The exact result is a probability that is
\emph{conditioned} on the placement.

@@

@@

\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\else
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
\fi
Each shaded node represents a beta-strand position.
Nodes connected by dashed edges are hydrogen-bonded.

@@

\begin{figure}%%%[h!]
\centerline{\includegraphics{speedup}}
\bigskip
\caption{Parallel speedup}
\label{fig:speedup}
\end{figure}

>>>>

\category{D.1.1}{Applicative (Functional) Programming}{}
\category{J.3}{Biology and genetics}{}
% \category{CR-number}{subcategory}{third-level}
% 
% \terms
% term1, term2
% 
\keywords
memoization, stochastic search, parallel strategies,  QuickCheck,
remote homology detection 
{\raggedright\par}

@@

To~account for beta strands, computational biologists extend
Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
\ifpagetuning
depends on the amino acid~$x'$ in column~${\pairedwith i}$.
\else
depends on whatever amino acid~$x'$  is in column~${\pairedwith i}$.
\fi
If~$x'$ is in position~$i'$ in the query sequence, Viterbi's
equations are altered so that, for example
$V_{j}^{\prime M}(i)$ depends not only on
$V_{j-1}^{\prime M}(i-1)$ but also on
$V_{\pairedwith j}^{\prime M}(i')$.\remark{I can't get an equation I
believe in. ---NR I think it's fine, unless we want to introduce
notation to distinguish paired amino acids from paired
model nodes --ND}
The distance between $j$~and~$\pairedwith j$ can be as small as a few
columns or as large as a few hundreds of columns.
Because $V_j^{\prime M}(i)$~depends not only on nearby values but also on
$V_{\pairedwith j}^{\prime M}(i')$,
dynamic programming cannot compute the maximum likelihood quickly 
\cite{Menke:2010ti,Daniels:2012}.

The change in Viterbi's equations is accompanied by a new model.
Within a beta strand, amino acids are not inserted or deleted, so
a bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}.

MRFy treats the beta 
strands in the model
 as ``beads'' which can slide along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's likelihood is computed
based on frequencies of amino-acid pairs observed in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the rest of
the query sequence, between and around beta strands,
is computed quickly and exactly 
using Viterbi's algorithm.
\ifpagetuning
This likelihood is \emph{conditioned} on the placement.
\else
The exact result is a probability that is
\emph{conditioned} on the placement.
\fi

@@

\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=7.8cm]{mrf_interleave_diagram.pdf}} 
\else
\centerline{\includegraphics[width=7.8cm]{mrf_interleave_diagram.eps}} 
\fi
Each shaded node represents a beta-strand position.
Nodes connected by dashed edges are hydrogen-bonded.

@@

@@

\begin{figure}%%%[h!]
\centerline{\includegraphics{speedup}}
\medskip
\caption{Parallel speedup}
\label{fig:speedup}
\end{figure}
